Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-14 concl.tex
forgot a few minor changes

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of an empirical
time series.  To operationalize that definition, we use an
approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
using a weighted version of the permutation entropy
of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
of forward information transfer in a time series---is ideal for our
purposes because it works with real-valued data and is known to
converge to the true entropy value. Using a variety of forecast models
and more than a hundred time-series data sets from a
computer-performance experiment, we have shown that prediction
accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  The
relationship is roughly logarithmic, which makes sense because any
signal with zero entropy should be perfectly predictable, and so
should ideally have a MASE value of zero.  Supporting our hypothesis,
forecast errors for the \col and \svdfive signals are higher than the
corresponding WPEs suggest---except for the nonlinear LMA predictor.
There are some exceptions: \svdone is easier to predict than its WPE
value suggests using any of the three forecast strategies studied
here.  On the whole, though, signals with higher weighted permutation
entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The curve in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
paragraph.  It is a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: a MASE score that is significantly higher than [[$3.58 \cdot
    \mathrm{WPE} - 1.91$]], where $\mathrm{WPE}$ is that of the
corresponding signal, is an indication that that time series has more
predictive structure than that forecast model can capture and exploit.

@@

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  Since these
patterns (dynamical chaos, periodicities, noise, etc.) are generic, we
believe that our results will generalize beyond this particular
system.  Testing that claim with different examples is another
important next step---and one that will be critical to validating our
claims about the utility of the dashed line as a heuristic for
determining whether a particular forecast method is well-matched to a
particular task.  Of particular interest would be the class of
so-called \emph{hybrid systems}~\cite{hybrid}, which exhibit discrete transitions
between different continuous regimes---e.g., a lathe that has an
intermittent instability or traffic at an internet router, whose
characteristic normal traffic patterns shift radically during an
attack.  Effective modeling and prediction of these kinds of systems
is quite difficult; doing so adaptively and automatically---in the
manner that is alluded to at the end of the previous paragraph---would
be an even more important challenge.

>>>>

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of an empirical
time series.  To operationalize that definition, we use an
approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
using a weighted version of the permutation entropy
of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
of forward information transfer in a time series---is ideal for our
purposes because it works with real-valued data and is known to
converge to the true entropy value. Using a variety of forecast models
and more than a hundred time-series data sets from a
computer-performance experiment, we have shown that prediction
accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  The
relationship is roughly logarithmic, which makes sense because any
signal with zero entropy should be perfectly predictable, and so
should ideally have a MASE value of zero.  [[Are there other reasons
    this shape makes sense?]]  Supporting our hypothesis, forecast
errors for the \col and \svdfive signals are higher than the
corresponding WPEs suggest---except for the nonlinear LMA predictor.
There are some exceptions: \svdone is easier to predict than its WPE
value suggests using any of the three forecast strategies studied
here.  On the whole, though, signals with higher weighted permutation
entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed curve and shaded region in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
operationalize the discussion in this paragraph.  These geometric
features are a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: [[a MASE score that is significantly higher than $3.58 \cdot
    \mathrm{WPE} - 1.91$, where $\mathrm{WPE}$ is that of the
    corresponding signal]] [[need to reword]], is an indication that
that time series has more predictive structure than that forecast
model can capture and exploit.

@@

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  Since these
patterns (dynamical chaos, periodicities, noise, etc.) are generic, we
believe that our results will generalize beyond this particular
system.  Testing that claim with different examples is another
important next step---and one that will be critical to validating our
claims about the utility of the shaded region around the dashed curve
in Figure [[what]] as a heuristic for determining whether a particular
forecast method is well-matched to a particular task.  Of particular
interest would be the class of so-called \emph{hybrid
  systems}~\cite{hybrid}, which exhibit discrete transitions between
different continuous regimes---e.g., a lathe that has an intermittent
instability or traffic at an internet router, whose characteristic
normal traffic patterns shift radically during an attack.  Effective
modeling and prediction of these kinds of systems is quite difficult;
doing so adaptively and automatically---in the manner that is alluded
to at the end of the previous paragraph---would be an even more
important challenge.

### 2014-07-14 pecomplexity.tex
forgot a few minor changes

\section*{Acknowledgment}
This work was partially supported by NSF grant \#CMMI-1245947 and ARO
grant \#W911NF-12-1-0288.
We would also like to acknowledge the Santa Fe Institute whose interdisciplinary halls created a place for this work to be born.

>>>>

\section*{Acknowledgment}
This work was partially supported by NSF grant \#CMMI-1245947 and ARO
grant \#W911NF-12-1-0288.  We would also like to acknowledge the Santa
Fe Institute whose interdisciplinary halls created a place for this
work to be born, and the anonymous reviewers whose comments
strengthened the presentation.

### 2014-07-14 results.tex
forgot a few minor changes

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values [[for all of these trials.]]
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_autolog}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of [[all trials for each system]].
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares log fit of all the points except
those from {\tt dgesdd$_1$}, which we have excluded for reasons
explained in the text.  [[Say what the colored region is and why we
    included it.]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the [[best-case]] prediction error for each of
these eight systems is roughly proportional to the log of the weighted
permutation entropy, which is consistent with our first conjecture.
The dashed line in the figure, a least-squares fit of $y = a \log(b x
+ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
  for reasons described below} captures this rough
proportionality\footnote{The logarithmic fit is appropriate here
  because any signal with zero entropy should be perfectly
  predictable, and so should ideally have a MASE value of zero.}.
This is not a formal result.  The forecast methods and data sets used
here were chosen to span the space of standard prediction strategies
and the range of dynamical behaviors, but they do not cover those
spaces exhaustively.  Our goal here is an\emph{empirical} assessment
of the relationship between predictability and complexity, not formal
results about a ``best'' predictor for a given time series.  There may
be other methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  [[(The
    points in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of
    the points for the corresponding system in any of the three plots
    in Figure~\ref{fig:wpe_vs_mase_all}.)]]  The WPE values do not
vary very much across trials.  For most traces, the variance in MASE
scores is low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
variance is larger, which spreads out the clusters horizontally.  The
distribution of the cluster \emph{positions} is quite different in the
three plots.  The mean MASE scores of predictions generated with the
nonlinear LMA method are generally closer to the dashed curve; the
ARIMA method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built \cite{autoARIMA}.  If a
KPSS test of the time series in question indicates that it is
nonstationary, an integration term is added to the model.  This test
gives mixed results in the case of the \col process, flagging five of
the 15 trials as stationary and ten as nonstationary.  ARIMA models
without an integration term perform more poorly on this signal, which
increases the error, thereby spreading out the points\footnote{We
  tested this hypothesis by forcing the inclusion of an integration
  term in the five cases where a KPSS test indicated that such a term
  was not needed.  This action removed the spread, pushing all 15 of
  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
  tight cluster.}.  \alert{This paragraph probably needs more, in
  response to the ref's gritch about fitting.}

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be leveraged to predict the future
course of the time series.  As mentioned above, though, the
effectiveness of any prediction strategy depends on how well it
leverages the information that is available in the data: methods that
use linear models, for instance, may fail when applied to nonlinear
processes.
%
% (This is the basis for the second conjecture above.)
%
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that there is more structure
in this signal than the corresponding method is able to leverage, and
that a different method might do better.  Indeed, for \svdfive, the
LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
of results that is much closer to the dashed curve on the WPE-MASE
plot.  Again, this validates our second conjecture: the LMA method can
capture and reproduce the way in which the \svdfive system processes
information, but the \naive ~and ARIMA prediction methods cannot.

@@

[[Short paragraph about how MASE has its drawbacks but is the best
    overall choice here because it is scale free and can compare
    across signals.  Point to the MASE paper for extensive evaluation
    and comparison wrt other metrics. We are in the process of
    repeating this study with other metrics, other forecast methods,
    and time-series data from other systems.  Segue to next section.]]

>>>>

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values [[for all of these trials.]]
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_autolog}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of [[all trials for each system]].
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares log fit of all the points except
those from {\tt dgesdd$_1$}, which we have excluded for reasons
explained in the text.  [[Say what the colored region is and why we
    included it.]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the [[best-case]] prediction error for each of
these eight systems is roughly proportional to the log of the weighted
permutation entropy, which is consistent with our first conjecture.
The dashed line in the figure, a least-squares fit of $y = a \log(b x
+ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
  for reasons described below} captures this relationship between WPE
and [[best-case]] prediction error\footnote{The logarithmic fit is
  appropriate here because any signal with zero entropy should be
  perfectly predictable, and so should ideally have a MASE value of
  zero.}.  This is not a formal result.  The forecast methods and data
sets used here were chosen to span the space of standard prediction
strategies and the range of dynamical behaviors, but they do not cover
those spaces exhaustively.  Our goal here is an\emph{empirical}
assessment of the relationship between predictability and complexity,
not formal results about a ``best'' predictor for a given time series.
There may be other methods that produce lower MASE values than those
in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
    shaded regions?  check in all captions]] is the same as in the
previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  [[(The
    points in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of
    the points for the corresponding system in any of the three plots
    in Figure~\ref{fig:wpe_vs_mase_all}.)]]  The WPE values do not
vary very much across trials.  For most traces, the variance in MASE
scores is low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
variance is larger, which spreads out the clusters horizontally.  The
distribution of the cluster \emph{positions} is quite different in the
three plots.  The mean MASE scores of predictions generated with the
nonlinear LMA method are generally closer to the dashed curve; the
ARIMA method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built \cite{autoARIMA}.  If a
KPSS test of the time series in question indicates that it is
nonstationary, the ARIMA recipe adds an integration term to the model.
This test gives mixed results in the case of the \col process,
flagging five of the 15 trials as stationary and ten as nonstationary.
ARIMA models without an integration term perform more poorly on this
signal, which increases the error, thereby spreading out the
points\footnote{We tested this hypothesis by forcing the inclusion of
  an integration term in the five cases where a KPSS test indicated
  that such a term was not needed.  This action removed the spread,
  pushing all 15 of the \col ARIMA points in
  Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
\alert{This paragraph probably needs more, in response to the ref's
  gritch about fitting.}

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be used to predict the future course of
the time series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the information
that is available in the data: methods that use linear models, for
instance, may fail when applied to nonlinear processes.
%
% (This is the basis for the second conjecture above.)
%
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that there is more structure
in this signal than the corresponding method is able to leverage, and
that a different method might do better.  Indeed, for \svdfive, the
LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
of results that is much closer to the dashed curve on the WPE-MASE
plot.  Again, this validates our second conjecture: the LMA method can
capture and reproduce the way in which the \svdfive system processes
information, but the \naive ~and ARIMA prediction methods cannot.

@@

[[Even though MASE has weaknesses, we concur with \cite{mase} that it
    is an effective way to compare prediction error across time series
    of different lengths and different scales.  Point to that paper
    for extensive evaluation and comparison wrt other metrics. We are
    in the process of repeating our own study with other metrics,
    other forecast methods, and time-series data from other systems.
    Segue to next section.]]
